Embedded Domain Specific Languages inside Functional Languages

### 2014-03-14 paper.tex
At end, and 4665 words.

Table~\cite{tab:status} gives the

The aim is to construct a structure, then
use that structure that was previously implicit.
Figure summarizes the state of the art with
deeply embedded reification

\begin{verbatim}
Expressions             Natural Fit
Functions               The trick
local bindings          Yes, using observable sharing
internal functions      Yes, using extension of OS
do-statements           Yes, using monadic reification
control flow            Not directly, can simulate
\end{verbatim}

The Deep EDSL trick breaks down on control flow,
which is not constructive, but instead requires
directly

\subsection{How to spot an imperative}

Sometimes, a list of actions is the best way to express what you want to do.
In Haskell, the \verb|do|-notation is used to chain together imperative actions.
To give an example,

\section{Reification Tricks}

The obvious questions are do this tricks used here scale,
and what are the limitations? The trick has
been used on may large examples, including
generating large telemerty circuits, 
and (( .. feldspar )). It is considered a
stardard pattern in the functional programming
community, analogoues to the OO patterns
enumerated in the gang of four book~\cite{..}.
The trick is made all the more appealing
because we can use operator overloading.
But the trick does has serious limitations.

The process of extracting a structure from a deeply embedded ESDL
is called reification.

 * running the description to get the AST

\section{Use of Haskell’s type system for various things }

 * expression data type 
 * the phantom types trick 
 * Use of type classes (e.g. distinction of Seq and Comb in Signal) 
 * usefulness of laziness 
 * smart constructors ? 
 * sharing ?

%--------------------------------
% CAREER12

\section{Domain Specific Languages}

We can also {\em reify\/} the circuit, into
a VHDL program. This is the key advantage and the
primary purpose of Kansas Lava. For example,
if we reify the counter with a width of 4 bits, we generate
the idiomatic but useable VHDL
listed in Figure~\ref{fig:counter-picture}.

This use of a Haskell DSL for expressing circuits
is all well and good and {\em structural\/}.
But how can Kansas Lava be utilized to
describe more general forms of computation and parallelism?
Unfortunately at this point the model breaks down
to some degree. Programs can be written
that generate specific components, like ALUs,
but computation is more challenging to express
in Kansas Lava and other deep DSLs.

>>>>

There are also extensions to the basic techniques. The principle once are:
\begin{itemize}
\item Internal function calls can captured as notes on a graph, 
rather than directly inlined~\cite{...}. 
This helps compilation of large programs, giving a basic separate compilation capability.
\item \verb|do|-statement can be reified, by normalization~\cite{...}. This result, called monadic
reification, is surprising. There are strong technical reasons to be believe monadic reification
should be impossible. However the normalization refactors the constraining that, by themselves would
be impossible to solve, and matches them up, 1-on-1, with a matching solution, allowing the whole
\verb|do|-notation to be solved and reified.
Monadic reification is a recent discovery, but has already been used in several Deep DSLs,
including feldspar, and Sunroof.
\item Control flow is problematic, and can not be used directly. However,
there is a generalization of Haskell boolean that do allow deep embedding capture~\cite{boolean}.
Using this library, a DSL with control flow can be constructed, but it needs to be explicit
code, at the DSL level, using constructors.
\end{itemize}

So where does this leave Deep DSLs? They are clearly a useful tool for the
language implementor, but come with costs and limitations. How can we therefore
push the state of the art, and allow more of the Haskell language to reified?
There are two primary shortcoming.
One we have discussed already: control flow and pattern matching remain a thorn in Deep DSLs.

Parametric polymorphism, one of the strengths of functional program, is the other
issue for deep DSLs. We need a specific structure to represent what we have captured,
and arbitrary polymorphism interferes with this. Current system sidestep this
by always instantiate at a specific type, but this expensive because size of
the captured program can expand exponentially.
Polymorphism was the technical reason that it was thought that monadic reification
would not work, but in that case it was side-stepped by normalization; this technique
does not generalize to all polymorphism.

Deep DSLs is the value-level way of extracting an expression, but there are others.
Quaziquoting is another, at the syntactically level. Haskell comes with an extensive
template system, called Template Haskell, that can and is used for DSLs.
(....)
Perhaps the future of Deep DSLs is some hybrid between expression generation
and quazi-quoting, combining the best of both systems.
